---
title: "Chapter 2"
output: html_notebook
---

# Chapter 2 R basics

In this book, we will be using the R software environment for all our analysis. You will learn R and data analysis techniques simultaneously. To follow along you will therefore need access to R. We also recommend the use of an integrated development environment (IDE), such as RStudio, to save your work. Note that it is common for a course or workshop to offer access to an R environment and an IDE through your web browser, as done by RStudio cloud12. If you have access to such a resource, you don’t need to install R and RStudio. However, if you intend on becoming an advanced data analyst, we highly recommend installing these tools on your computer13. Both R and RStudio are free and available online.


##2.1 Case study: US Gun Murders


Imagine you live in Europe and are offered a job in a US company with many locations across all states. It is a great job, but news with headlines such as US Gun Homicide Rate Higher Than Other Developed Countries14 have you worried. Charts like this may concern you even more:

Or even worse, this version from everytown.org:

But then you remember that the US is a large and diverse country with 50 very different states as well as the District of Columbia (DC).

California, for example, has a larger population than Canada, and 20 US states have populations larger than that of Norway. In some respects, the variability across states in the US is akin to the variability across countries in Europe. Furthermore, although not included in the charts above, the murder rates in Lithuania, Ukraine, and Russia are higher than 4 per 100,000. So perhaps the news reports that worried you are too superficial. You have options of where to live and want to determine the safety of each particular state. We will gain some insights by examining data related to gun homicides in the US during 2010 using R.

Before we get started with our example, we need to cover logistics as well as some of the very basic building blocks that are required to gain more advanced R skills. Be aware that the usefulness of some of these building blocks may not be immediately obvious, but later in the book you will appreciate having mastered these skills.

##2.2 The very basics
Before we get started with the motivating dataset, we need to cover the very basics of R.

```{r}
A <- 1
A = 1
```
```{r}
B = 'apple'
C = 'apple'
```


###2.2.1 Objects
Suppose a high school student asks us for help solving several quadratic equations of the form  
$ax^2+bx+c=0$

 . The quadratic formula gives us the solutions:

−
b
−
√
b
2
−
4
a
c
2
a
 and 
−
b
+
√
b
2
−
4
a
c
2
a
 
which of course change depending on the values of  
a
 ,  
b
 , and  
c
 . One advantage of programming languages is that we can define variables and write expressions with these variables, similar to how we do so in math, but obtain a numeric solution. We will write out general code for the quadratic equation below, but if we are asked to solve  
x
2
+
x
−
1
=
0
 , then we define:


```{r}
a <- 1
b <- 1
c <- -1
```


which stores the values for later use. We use <- to assign values to the variables.

We can also assign values using = instead of <-, but we recommend against using = to avoid confusion.

Copy and paste the code above into your console to define the three variables. Note that R does not print anything when we make this assignment. This means the objects were defined successfully. Had you made a mistake, you would have received an error message.

To see the value stored in a variable, we simply ask R to evaluate a and it shows the stored value:
```{r}
a
```

A more explicit way to ask R to show us the value stored in a is using print like this:
```{r}
print(a)
```
We use the term object to describe stuff that is stored in R. Variables are examples, but objects can also be more complicated entities such as functions, which are described later.

###2.2.2 The workspace
As we define objects in the console, we are actually changing the workspace. You can see all the variables saved in your workspace by typing:
```{r}
ls()
```

In RStudio, the Environment tab shows the values:



We should see a, b, and c. If you try to recover the value of a variable that is not in your workspace, you receive an error. For example, if you type x you will receive the following message: Error: object 'x' not found.

Now since these values are saved in variables, to obtain a solution to our equation, we use the quadratic formula:
```{r}
(-b + sqrt(b^2 - 4*a*c) ) / ( 2*a )
```


```{r}
(-b - sqrt(b^2 - 4*a*c) ) / ( 2*a )
```


###2.2.3 Functions
Once you define variables, the data analysis process can usually be described as a series of functions applied to the data. R includes several predefined functions and most of the analysis pipelines we construct make extensive use of these.

We already used the install.packages, library, and ls functions. We also used the function sqrt to solve the quadratic equation above. There are many more prebuilt functions and even more can be added through packages. These functions do not appear in the workspace because you did not define them, but they are available for immediate use.

In general, we need to use parentheses to evaluate a function. If you type ls, the function is not evaluated and instead R shows you the code that defines the function. If you type ls() the function is evaluated and, as seen above, we see objects in the workspace.

Unlike ls, most functions require one or more arguments. Below is an example of how we assign an object to the argument of the function log. Remember that we earlier defined a to be 1:
```{r}
log(8)
log(a) 
```

You can find out what the function expects and what it does by reviewing the very useful manuals included in R. You can get help by using the help function like this:
```{r}
help("log")
```
For most functions, we can also use this shorthand:
```{r}
?log
```

The help page will show you what arguments the function is expecting. For example, log needs x and base to run. However, some arguments are required and others are optional. You can determine which arguments are optional by noting in the help document that a default value is assigned with =. Defining these is optional. For example, the base of the function log defaults to base = exp(1) making log the natural log by default.

If you want a quick look at the arguments without opening the help system, you can type:
```{r}
args(log)
```
You can change the default values by simply assigning another object:
```{r}
log(8, base = 2)
```

Note that we have not been specifying the argument x as such:
```{r}
log(x = 8, base = 2)
```
The above code works, but we can save ourselves some typing: if no argument name is used, R assumes you are entering arguments in the order shown in the help file or by args. So by not using the names, it assumes the arguments are x followed by base:
```{r}
log(8,2)
```
If using the arguments’ names, then we can include them in whatever order we want:
```{r}
log(base = 2, x = 8)
```
To specify arguments, we must use =, and cannot use <-.

There are some exceptions to the rule that functions need the parentheses to be evaluated. Among these, the most commonly used are the arithmetic and relational operators. For example:
```{r}
2 ^ 3
```
You can see the arithmetic operators by typing:
```{r}
help("+")
```

or
```{r}
?"+"
```

and the relational operators by typing:
```{r}
help(">")
```

or
```{r}
?">"
```

###2.2.4 Other prebuilt objects
There are several datasets that are included for users to practice and test out functions. You can see all the available datasets by typing:
```{r}
data()
```
This shows you the object name for these datasets. These datasets are objects that can be used by simply typing the name. For example, if you type:
```{r}
co2
```

R will show you Mauna Loa atmospheric CO2 concentration data.

Other prebuilt objects are mathematical quantities, such as the constant  
π and ∞:
```{r}
pi

Inf+1

```
###2.2.5 Variable names
We have used the letters a, b, and c as variable names, but variable names can be almost anything. Some basic rules in R are that variable names have to start with a letter, can’t contain spaces, and should not be variables that are predefined in R. For example, don’t name one of your variables install.packages by typing something like install.packages <- 2.

A nice convention to follow is to use meaningful words that describe what is stored, use only lower case, and use underscores as a substitute for spaces. For the quadratic equations, we could use something like this:

```{r}
solution_1 <- (-b + sqrt(b^2 - 4*a*c)) / (2*a)
solution_2 <- (-b - sqrt(b^2 - 4*a*c)) / (2*a)
```

For more advice, we highly recommend studying Hadley Wickham’s style guide

###2.2.6 Saving your workspace
Values remain in the workspace until you end your session or erase them with the function rm. But workspaces also can be saved for later use. In fact, when you quit R, the program asks you if you want to save your workspace. If you do save it, the next time you start R, the program will restore the workspace.

We actually recommend against saving the workspace this way because, as you start working on different projects, it will become harder to keep track of what is saved. Instead, we recommend you assign the workspace a specific name. You can do this by using the function save or save.image. To load, use the function load. When saving a workspace, we recommend the suffix rda or RData. In RStudio, you can also do this by navigating to the Session tab and choosing Save Workspace as. You can later load it using the Load Workspace options in the same tab. You can read the help pages on save, save.image, and load to learn more.

###2.2.7 Motivating scripts
To solve another equation such as  
3x^2+2x−1, we can copy and paste the code above and then redefine the variables and recompute the solution:

```{r}
a <- 3
b <- 2
c <- -1
(-b + sqrt(b^2 - 4*a*c)) / (2*a)
(-b - sqrt(b^2 - 4*a*c)) / (2*a)
```
By creating and saving a script with the code above, we would not need to retype everything each time and, instead, simply change the variable names. Try writing the script above into an editor and notice how easy it is to change the variables and receive an answer.

###2.2.8 Commenting your code
If a line of R code starts with the symbol #, it is not evaluated. We can use this to write reminders of why we wrote particular code. For example, in the script above we could add:

```{r}
## Code to compute solution to quadratic equation of the form ax^2 + bx + c
## define the variables
a <- 3 
b <- 2
c <- -1

## now compute the solution
(-b + sqrt(b^2 - 4*a*c)) / (2*a)
(-b - sqrt(b^2 - 4*a*c)) / (2*a)
```
##2.3 Exercises
1. What is the sum of the first 100 positive integers? The formula for the sum of integers 1 through n is  n(n+1)/2. Define  n=100 and then use R to compute the sum of  1 through  100 using the formula. What is the sum?
```{r}
n <- 100
n*(n+1)/2
```

2. Now use the same formula to compute the sum of the integers from 1 through 1,000.
```{r}
n <- 1000
n*(n+1)/2
```

3. Look at the result of typing the following code into R:
```{r}
n <- 1000
x <- seq(1, n)
sum(x)
```

Based on the result, what do you think the functions seq and sum do? You can use help.

b

4. In math and programming, we say that we evaluate a function when we replace the argument with a given number. So if we type sqrt(4), we evaluate the sqrt function. In R, you can evaluate a function inside another function. The evaluations happen from the inside out. Use one line of code to compute the log, in base 10, of the square root of 100.


```{r}
log10(sqrt(100))
```

5. Which of the following will always return the numeric value stored in x? You can try out examples and use the help system if you want.
```{r}
log(10^x)
log10(x^10)
log(exp(x))
exp(log(x, base = 2))

```

log(10^x)
log10(x^10)
log(exp(x))
exp(log(x, base = 2))


ans c

##2.4 Data types
Variables in R can be of different types. For example, we need to distinguish numbers from character strings and tables from simple lists of numbers. The function class helps us determine what type of object we have:
```{r}
a <- 2
class(a)

```

###2.4.1 Data frames
Up to now, the variables we have defined are just one number. This is not very useful for storing data. The most common way of storing a dataset in R is in a data frame. Conceptually, we can think of a data frame as a table with rows representing observations and the different variables reported for each observation defining the columns. Data frames are particularly useful for datasets because we can combine different data types into one object.

A large proportion of data analysis challenges start with data stored in a data frame. For example, we stored the data for our motivating example in a data frame. You can access this dataset by loading the dslabs library and loading the murders dataset using the data function:
```{r}
library(dslabs)
data(murders)
```

To see that this is in fact a data frame, we type:
```{r}
class(murders)
```

###2.4.2 Examining an object
The function str is useful for finding out more about the structure of an object:
```{r}
str(murders)
```

This tells us much more about the object. We see that the table has 51 rows (50 states plus DC) and five variables. We can show the first six lines using the function head:
```{r}
head(murders)
```
In this dataset, each state is considered an observation and five variables are reported for each state.

Before we go any further in answering our original question about different states, let’s learn more about the components of this object.

###2.4.3 The accessor: $
For our analysis, we will need to access the different variables represented by columns included in this data frame. To do this, we use the accessor operator $ in the following way:


```{r}
murders$population
```

But how did we know to use population? Previously, by applying the function str to the object murders, we revealed the names for each of the five variables stored in this table. We can quickly access the variable names using:
```{r}
names(murders)
```

It is important to know that the order of the entries in murders$population preserves the order of the rows in our data table. This will later permit us to manipulate one variable based on the results of another. For example, we will be able to order the state names by the number of murders.

Tip: R comes with a very nice auto-complete functionality that saves us the trouble of typing out all the names. Try typing murders$p then hitting the tab key on your keyboard. This functionality and many other useful auto-complete features are available when working in RStudio.

###2.4.4 Vectors: numerics, characters, and logical
The object murders$population is not one number but several. We call these types of objects vectors. A single number is technically a vector of length 1, but in general we use the term vectors to refer to objects with several entries. The function length tells you how many entries are in the vector:
```{r}
pop <- murders$population
length(pop)
```

This particular vector is numeric since population sizes are numbers:
```{r}
class(pop)
```
In a numeric vector, every entry must be a number.

To store character strings, vectors can also be of class character. For example, the state names are characters:
```{r}
class(murders$state)
```

As with numeric vectors, all entries in a character vector need to be a character.

Another important type of vectors are logical vectors. These must be either TRUE or FALSE.
```{r}
z <- 3 == 2
z
class(z)

```
Here the == is a relational operator asking if 3 is equal to 2. In R, if you just use one =, you actually assign a variable, but if you use two == you test for equality.

You can see the other relational operators by typing:
```{r}
?Comparison
```

In future sections, you will see how useful relational operators can be.

We discuss more important features of vectors after the next set of exercises.

Advanced: Mathematically, the values in pop are integers and there is an integer class in R. However, by default, numbers are assigned class numeric even when they are round integers. For example, class(1) returns numeric. You can turn them into class integer with the as.integer() function or by adding an L like this: 1L. Note the class by typing: class(1L)

###2.4.5 Factors
In the murders dataset, we might expect the region to also be a character vector. However, it is not:
```{r}
class(murders$region)
```
It is a factor. Factors are useful for storing categorical data. We can see that there are only 4 regions by using the levels function:
```{r}
levels(murders$region)
```
In the background, R stores these levels as integers and keeps a map to keep track of the labels. This is more memory efficient than storing all the characters.

Note that the levels have an order that is different from the order of appearance in the factor object. The default in R is for the levels to follow alphabetical order. However, often we want the levels to follow a different order. You can specify an order through the levels argument when creating the factor with the factor function. For example, in the murders dataset regions are ordered from east to west. The function reorder lets us change the order of the levels of a factor variable based on a summary computed on a numeric vector. We will demonstrate this with a simple example, and will see more advanced ones in the Data Visualization part of the book.

Suppose we want the levels of the region by the total number of murders rather than alphabetical order. If there are values associated with each level, we can use the reorder and specify a data summary to determine the order. The following code takes the sum of the total murders in each region, and reorders the factor following these sums.

```{r}
region <- murders$region
value <- murders$total
region <- reorder(region, value, FUN = sum)
levels(region)
```

The new order is in agreement with the fact that the Northeast has the least murders and the South has the most.

Warning: Factors can be a source of confusion since sometimes they behave like characters and sometimes they do not. As a result, confusing factors and characters are a common source of bugs.

###2.4.6 Lists
Data frames are a special case of lists. Lists are useful because you can store any combination of different types. You can create a list using the list function like this:

```{r}
record <- list(name = "John Doe",
             student_id = 1234,
             grades = c(95, 82, 91, 97, 93),
             final_grade = "A")
```

The function c is described in Section 2.6.

This list includes a character, a number, a vector with five numbers, and another character.
```{r}
record
```

As with data frames, you can extract the components of a list with the accessor $.
```{r}
record$student_id
```
We can also use double square brackets ([[) like this:

```{r}
record[["student_id"]]
```
You should get used to the fact that in R, there are often several ways to do the same thing, such as accessing entries.

You might also encounter lists without variable names.
```{r}
record2 <- list("John Doe", 1234)
record2
```
If a list does not have names, you cannot extract the elements with $, but you can still use the brackets method and instead of providing the variable name, you provide the list index, like this:
```{r}
record2[[1]]
```
We won’t be using lists until later, but you might encounter one in your own exploration of R. For this reason, we show you some basics here.

###2.4.7 Matrices
Matrices are another type of object that are common in R. Matrices are similar to data frames in that they are two-dimensional: they have rows and columns. However, like numeric, character and logical vectors, entries in matrices have to be all the same type. For this reason data frames are much more useful for storing data, since we can have characters, factors, and numbers in them.

Yet matrices have a major advantage over data frames: we can perform matrix algebra operations, a powerful type of mathematical technique. We do not describe these operations in this book, but much of what happens in the background when you perform a data analysis involves matrices. We cover matrices in more detail in Chapter 33.1 but describe them briefly here since some of the functions we will learn return matrices.

We can define a matrix using the matrix function. We need to specify the number of rows and columns.
```{r}
mat <- matrix(1:12, 4, 3)
mat
```
You can access specific entries in a matrix using square brackets ([). If you want the second row, third column, you use
```{r}
mat[2, 3]
```

If you want the entire second row, you leave the column spot empty:
```{r}
mat[2, ]
```
Notice that this returns a vector, not a matrix.

Similarly, if you want the entire third column, you leave the row spot empty:
```{r}
mat[, 3]
```

This is also a vector, not a matrix.

You can access more than one column or more than one row if you like. This will give you a new matrix.
```{r}
mat[, 2:3]
```

You can subset both rows and columns:
```{r}
mat[1:2, 2:3]

```
We can convert matrices into data frames using the function as.data.frame:
```{r}
as.data.frame(mat)
```
You can also use single square brackets ([) to access rows and columns of a data frame:

```{r}
data("murders")
murders[25, 1]


```

```{r}
murders[2:3, ]
```

##2.5 Exercises
1. Load the US murders dataset.
```{r}
library(dslabs)
data(murders)
```


Use the function str to examine the structure of the murders object. Which of the following best describes the variables represented in this data frame?

```{r}
str(murders)
```
a

2. What are the column names used by the data frame for these five variables?

state, abb, region, population, total

3. Use the accessor $ to extract the state abbreviations and assign them to the object a. What is the class of this object?
```{r}
a<- murders$abb
class(a)
```



4. Now use the square brackets to extract the state abbreviations and assign them to the object b. Use the identical function to determine if a and b are the same.
```{r}
b <-murders [["abb"]]
identical(a,b)
```

5. We saw that the region column stores a factor. You can corroborate this by typing:

```{r}
class(murders$region)
```
```{r}
length(levels(murders$region))
```
With one line of code, use the function levels and length to determine the number of regions defined by this dataset.

6. The function table takes a vector and returns the frequency of each element. You can quickly see how many states are in each region by applying this function. Use this function in one line of code to create a table of states per region.
```{r}
table(murders$region)
```

##2.6 Vectors
In R, the most basic objects available to store data are vectors. As we have seen, complex datasets can usually be broken down into components that are vectors. For example, in a data frame, each column is a vector. Here we learn more about this important class.

###2.6.1 Creating vectors
We can create vectors using the function c, which stands for concatenate. We use c to concatenate entries in the following way:
```{r}
codes <- c(380, 124, 818)
codes
```

We can also create character vectors. We use the quotes to denote that the entries are characters rather than variable names.
```{r}
country <- c("italy", "canada", "egypt")
```

In R you can also use single quotes:

```{r}
country <- c('italy', 'canada', 'egypt')
```
But be careful not to confuse the single quote ’ with the back quote `.

By now you should know that if you type:
```{r}
country <- c(italy, canada, egypt)
```

you receive an error because the variables italy, canada, and egypt are not defined. If we do not use the quotes, R looks for variables with those names and returns an error.

###2.6.2 Names
Sometimes it is useful to name the entries of a vector. For example, when defining a vector of country codes, we can use the names to connect the two:

```{r}
codes <- c(italy = 380, canada = 124, egypt = 818)
codes
```
The object codes continues to be a numeric vector:


```{r}
class(codes)
```
but with names:
```{r}
names(codes)
```

If the use of strings without quotes looks confusing, know that you can use the quotes as well:
```{r}
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)
codes
```
There is no difference between this function call and the previous one. This is one of the many ways in which R is quirky compared to other languages.

We can also assign names using the names functions:
```{r}
codes <- c(380, 124, 818)
country <- c("italy","canada","egypt")
names(codes) <- country
codes
```
###2.6.3 Sequences
Another useful function for creating vectors generates sequences:

```{r}
seq(1, 10)

```
The first argument defines the start, and the second defines the end which is included. The default is to go up in increments of 1, but a third argument lets us tell it how much to jump by:

```{r}
seq(1, 10, 2)
```
If we want consecutive integers, we can use the following shorthand:
```{r}
1:10
```
When we use these functions, R produces integers, not numerics, because they are typically used to index something:
```{r}
class(1:10)
```
However, if we create a sequence including non-integers, the class changes:
```{r}
class(seq(1, 10, 0.5))
```
###2.6.4 Subsetting
We use square brackets to access specific elements of a vector. For the vector codes we defined above, we can access the second element using:
```{r}
codes[2]
```
You can get more than one entry by using a multi-entry vector as an index:
```{r}
codes[c(1,3)]
```
The sequences defined above are particularly useful if we want to access, say, the first two elements:
```{r}
codes[1:2]

```
If the elements have names, we can also access the entries using these names. Below are two examples.

```{r}
codes["canada"]
```








Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
